---
title: "dbGaPCheckup"
output: 
  github_document:
    number_sections: true
# To update the *.md version of this, run devtools::build_readme()
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# Overview

<!-- badges: start -->
<!-- badges: end -->

*I want to make the data sing, but it is torturing me instead. Will the real data please stand up?*

The goal of `dbGaPCheckup` is to make your National Library of Medicine database of Genotypes and Phenotypes ([dbGaP](https://www.ncbi.nlm.nih.gov/gap/)) data set submission a tiny bit easier. Specifically, our package implements several check, awareness, utility, and reporting functions designed to help you ensure that your Subject Phenotype data set and data dictionary meet a variety of dbGaP specific [formatting requirements](https://www.ncbi.nlm.nih.gov/gap/docs/submissionguide/). A list of the functions available can be found below. 

The software announcement for our package has been published in BMC Bioinformatics and is available at https://bmcbioinformatics.biomedcentral.com/articles/10.1186/s12859-023-05200-8. 

Heinsberg, L.W., Weeks, D.E. dbGaPCheckup: pre-submission checks of dbGaP-formatted subject phenotype files. BMC Bioinformatics 24, 77 (2023). https://doi.org/10.1186/s12859-023-05200-8

```{r individual_checks, echo=FALSE}
fn.path <- system.file("extdata", "Functions.xlsx",
   package = "dbGaPCheckup", mustWork=TRUE)
fns <- readxl::read_xlsx(fn.path)
knitr::kable(fns, caption="List of function names and types.")
```

# Copyright information 

Copyright 2022, University of Pittsburgh. All Rights Reserved.
License: [GPL-2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)

# Installation

You can install dbGaPCheckup from [CRAN](https://cran.r-project.org/) using: 

```
install.packages("dbGaPCheckup")
```

You can install the development version of dbGaPCheckup from [GitHub](https://github.com/) using:

``` r
# install.packages("devtools")
devtools::install_github("lwheinsberg/dbGaPCheckup")
```

# Example

An introductory example is provided below. For more details see https://lwheinsberg.github.io/dbGaPCheckup/ or the `dbGaPCheckup_vignette` (expanded instructions) or `dbGaPCheckup` (Quick Start) vignette documents.

A special note: As you will see below, this package requires several fields beyond those required by the dbGaP [formatting requirements](https://www.ncbi.nlm.nih.gov/gap/docs/submissionguide/). Specifically, while dbGaP requires that only the `VARNAME`, `VARDESC`, `UNITS`, and `VALUES` columns be present in the data dictionary, this package requires that `MIN`, `MAX`, and `TYPE` fields are also included. If your data dictionary does not include these fields already, you can use the `add_missing_fields` function to auto fill them (see below).

```{r library}
library(dbGaPCheckup)
```

Read in your Subject Phenotype data into `DS.data`.

```{r data}
DS.path <- system.file("extdata", "DS_Example.txt",
   package = "dbGaPCheckup", mustWork=TRUE)
DS.data <- read.table(DS.path, header=TRUE, sep="\t", quote="", as.is = TRUE)
```

Read in your Subject Phenotype data dictionary into `DD.dict`.

```{r dict}
DD.path <- system.file("extdata", "DD_Example2f.xlsx",
   package = "dbGaPCheckup", mustWork=TRUE)
DD.dict <- readxl::read_xlsx(DD.path)
```

Run the function `check_report`. Note that, for many functions, specification of missing value codes are important for accurate results. 

```{r cr1}
report <- check_report(DD.dict = DD.dict, DS.data = DS.data, non.NA.missing.codes=c(-4444, -9999))
```

As described in more detail in the `dbGaPCheckup_vignette` vignette, some checks contain embedded "pre-checks" that must be passed before the check can be run. For example,  as mentioned above, this package requires `MIN`, `MAX`, and `TYPE` fields in the data dictionary. We can see above that several of the checks (e.g., `NA_check`, `integer_check`, `decimal_check`) were not run because these additional required fields were missing. Never fear though! We have created a function to auto fill these fields that can be used to get further along in the checks!

```{r add_missing}
DD.dict.updated <- add_missing_fields(DD.dict, DS.data)
```

Once the fields are added, you can simply return to run your checks. Don't forget to use the updated dictionary though! 

```{r cr2}
report.v2 <- check_report(DD.dict = DD.dict.updated , DS.data = DS.data, non.NA.missing.codes=c(-4444, -9999))
```

Now we see that 13 out of 15 checks pass, but the workflow fails at `description_check` and `missing_value_check`. Specifically, in `description_check` we see that variables `PREGNANT` and `REACT` were identified as having missing variable descriptions (`VARDESC`), and variables `HEIGHT` and `WEIGHT` incorrectly have identical descriptions. In `missing_value_check`, we see that the variable `CUFFSIZE` contains a `-9999` encoded value that is not specified in a `VALUES` column. While we have included functions that support "simple fixes", the issues identified here would need to be corrected manually in your data dictionary before moving on. 

Note that our package also includes some helpful reporting functions (`create_report` and `create_awareness_report`) that can help you more fully interrogate your data and catch potential errors prior to dbGaP data submission.

Finally, after your data dictionary is fully consistent with your data, you can use the `label_data` function to convert your data to labelled data, essentially embedding the data dictionary into the data.  

For more information about these reporting functions, and all of the checks available in our package, see the Quick Start (`dbGaPCheckup`) or expanded (`dbGaPCheckup_vignette`) vignettes for more information! 

# Contact information 

If you have any questions or comments, please feel free to contact us!           

Lacey W. Heinsberg, PhD, RN: law145@pitt.edu             
Daniel E. Weeks, PhD: weeks@pitt.edu                 

Bug reports: https://github.com/lwheinsberg/dbGaPCheckup/issues

# Acknowledgments

This package was developed with partial support from the National Institutes of Health under award numbers R01HL093093, R01HL133040, and K99HD107030. The `eval_function` and `dat_function` functions that form the backbone of the awareness reports were inspired by an elegant 2016 homework answer submitted by Tanbin Rahman in our HUGEN 2070 course ‘Bioinformatics for Human Genetics’. We would also like to thank Nick Moshgat for testing and providing feedback on our package during development. 
