---
title: "dbGaPCheckup Quick Start"
author: "Lacey W. Heinsberg and Daniel E. Weeks\n"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  rmarkdown::html_vignette:
    toc: true
    toc_depth: 3
    number_sections: true
pkgdown:
  as_is: true 
  toc: true
  toc_depth: 3
  number_sections: true 
vignette: >
  %\VignetteIndexEntry{dbGaPCheckup Quick Start}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
# To render to GitHub markdown:
# render("dbGaPCheckup.Rmd",github_document(toc=TRUE, toc_depth=3, number_sections=TRUE))

---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Copyright information 

Copyright 2022, University of Pittsburgh. All Rights Reserved.
License: [GPL-2](https://www.gnu.org/licenses/old-licenses/gpl-2.0.en.html)

# Installation 

To install from [CRAN](https://cran.r-project.org/) use: 

```
install.packages("dbGaPCheckup")
```

To install the development version from [GitHub](https://github.com/) use: 

```
devtools::install_github("lwheinsberg/dbGaPCheckup/pkg")
```

# Quick start 

This document is designed to provide "quick start" guidance for using the `dbGaPCheckUp` R package. Please see the table below and `dbGaPCheckup_vignette` for more detailed information. 

```{r individual_checks, echo=FALSE}
fn.path <- system.file("extdata", "Functions.xlsx",
   package = "dbGaPCheckup", mustWork=TRUE)
fns <- readxl::read_xlsx(fn.path)
knitr::kable(fns, caption="List of function names and types.")
```

# Usage

After the `dbGaPCheckup` package has been installed, you can load the R package using this command: 

`library(dbGaPCheckup)`

Then proceed as follows:

1. Read in your data into `DS.data`; 
2. Read in your data dictionary into `DD.dict`; 
3. Run the function `check_report`, optionally defining any missing value codes (e.g., -9999) via the `non.NA.missing.codes` argument.

Note, as you will see below, this package requires several fields beyond those required by the dbGaP [formatting requirements](https://www.ncbi.nlm.nih.gov/gap/docs/submissionguide/). Specifically, the data dictionary is required to also have `MIN`, `MAX`, and `TYPE` fields. If your data dictionary does not include these fields already, you can use the `add_missing_fields` function to auto fill them (see below).

# Example usage

## Load the `dbGaPCheckup` R package

```{r load_libraries}
library(dbGaPCheckup)
```

##  Read in your Subject Phenotype data into `DS.data`.

```{r data}
DS.path <- system.file("extdata", "DS_Example.txt",
   package = "dbGaPCheckup", mustWork=TRUE)
DS.data <- read.table(DS.path, header=TRUE, sep="\t", quote="", as.is = TRUE)
```

## Read in your Subject Phenotype data dictionary into `DD.dict`.

```{r dict}
DD.path <- system.file("extdata", "3b_SSM_DD_Example2f.xlsx",
   package = "dbGaPCheckup", mustWork=TRUE)
DD.dict <- readxl::read_xlsx(DD.path)
```

## Run the function `check_report`.

With many functions, specification of missing value codes are important for accurate results. 

```{r cr1}
report <- check_report(DD.dict = DD.dict, DS.data = DS.data, non.NA.missing.codes=c(-4444, -9999))
```

As described in more detail in the `dbGaPCheckup_vignette` vignette, some checks contain embedded "pre-checks" that must be passed before the check can be run. For example,  As mentioned above, this package requires `MIN`, `MAX`, and `TYPE` fields in the data dictionary. We have created a function to auto fill these fields that can be used to get further along in the checks. 

```{r add_missing}
DD.dict.updated <- add_missing_fields(DD.dict, DS.data)
```

Once the fields are added, you can return to run your checks.

```{r cr2}
report.v2 <- check_report(DD.dict = DD.dict.updated , DS.data = DS.data, non.NA.missing.codes=c(-4444, -9999))
```

Now we see that 13 out of 15 checks pass, but the workflow fails at `description_check` and `missing_value_check`. Specifically, in `description_check` we see that variables `PREGNANT` and `REACT` were identified as having missing variable descriptions (`VARDESC`), and variables `HEIGHT` and `WEIGHT` incorrectly have identical descriptions. In `missing_value_check`, we see that the variable `CUFFSIZE` contains a `-9999` encoded value that is not specified in a `VALUES` column. While we have included functions that support "simple fixes", the issues identified here would need to be corrected manually in your data dictionary before moving on. 

## Reporting

Note that we have also created reporting functions that generate graphical and textual descriptions and awareness checks of the data in HTML format (see `dbGaPCheckup_vignette` vignette: `create_awareness_report` (Appendix A) and `create_report` (Appendix B) for more details). These reports are designed to help you catch other potential errors in your data set. Note that the `create_report` generated is quite long however, so we recommend that you only submit subsets of variables at a time. Specification of missing value codes are also important for effective plotting. 


```
# Functions not run here as they work best when initiated interactively
# Awareness Report (See Appendix A of the `dbGaPCheckup` vignette)
create_awareness_report(DD.dict.updated, DS.data, non.NA.missing.codes=c(-9999, -4444),
   output.path= tempdir())
   
# Data Report (See Appendix B of the `dbGaPCheckup` vignette)
create_report(DD.dict.updated, DS.data, sex.split=TRUE, sex.name= "SEX",
   start = 3, end = 7, non.NA.missing.codes=c(-9999,-4444),
   output.path= tempdir(), open.html=TRUE)
```

More details on execution and interpretation have been provided in the `dbGaPCheckup_vignette` vignette. 

# Labelled data

After your data dictionary is fully consistent with your data, you can use the `label_data` function to convert your data to labelled data, essentially embedding the data dictionary into the data for future use!  

```{r label_data}
DS_labelled_data <- label_data(DD.dict.updated, DS.data, non.NA.missing.codes=c(-9999))
labelled::var_label(DS_labelled_data$SEX)
labelled::val_labels(DS_labelled_data$SEX)
attributes(DS_labelled_data$SEX)
labelled::na_values(DS_labelled_data$HX_DEPRESSION)
```

# Contact information 

If you have any questions or comments, please feel free to contact us! 

Lacey W. Heinsberg: law145@pitt.edu       
Daniel E. Weeks: weeks@pitt.edu             

Bug reports: https://github.com/lwheinsberg/dbGaPCheckup/issues

# Acknowledgments

This package was developed with partial support from the National Institutes of Health under award numbers R01HL093093, R01HL133040, and K99HD107030. The `eval_function` and `dat_function` functions that form the backbone of the awareness reports were inspired by an elegant 2016 homework answer submitted by Tanbin Rahman in our HUGEN 2070 course ‘Bioinformatics for Human Genetics’. We would also like to thank Nick Moshgat for testing and providing feedback on our package during development.
